---
title: "R for Marine Science - Workshop 2"
author: "Katie Dyck"
date: "2026-06-04"
execute:
  echo: false
format:
    html:
        embed-resources: true
        code-fold: show
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

## Introduction

This qmd file includes all material covered from the second "R for Marine Science" workshop. The material covered includes:

- Tidying data
- Pivoting wider and longer
- Separating and uniting variables
- Dealing with strings and dates
- Joining and separating data frames/tibbles
- Handling missing values

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r load packages}
# Load all required packages for this workshop
library(tidyverse)  # includes `stringr` and `lubridate` packages
library(here)
library(readxl)
library(palmerpenguins)
```

# Tidying data

This is what a tidy data table looks like:

```{r tidy table example}
table1 <- table1
print(table1)
```

This is what an *untidy* data table looks like:

```{r untidy table examples}
table2 <- table2
print(table2)

table3 <- table3
print(table3)
```

A tidy dataset follows these 3 interrelated rules:

- Each variable has its own column
- Each observation has its own row
- Each value has its own cell

The rules are interrelated because only 2 of the 3 rules can be followed at any time.

*All `tidyverse` packages have been designed to work with tidy data.*

# Pivoting data

`pivot_longer()` and `pivot_wider()` are used to reshape data tables. Both functions are made available from the `tidyr` package, which is part of the `tidyverse`.

```{r pivot_longer()}
table1 |> 
  pivot_longer(cols = c("cases", "population"),
               names_to = "case_type",
               values_to = "case_count")
```
```{r pivot_wider}
table2 |> 
  pivot_wider(names_from = "type", values_from = "count")
```


## Lengthening datasets

`pivot_longer()` is used to lengthen datasets. It takes multiple columns and collapses them into key-value pairs, *increasing* the number of rows and *decreasing* the number of columns. The dataset is split by the `names_to` argument, which specifies the name of the new column that will contain the names of the original columns. The `values_to` argument specifies the name of the new column that will contain the values from the original columns.

```{r billboard}
billboard <- billboard
billboard
```

In this dataset, each observation is a song, and the first 3 columns (`artist`, `track`, and `date.entered`) are variables that describe the song. The remaining columns (`wk1` to `wk76`) are not variables, but rather values *of* a variable (week). To tidy this dataset so that each row is a single observation, we can use `pivot_longer()` to collapse the `wk1` to `wk76` columns into key-value pairs.

```{r billboard pivot_longer}
billboard |> 
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               values_to = "rank")
```

As some songs are only on the billboard for a few weeks, there are many `NA` values in the `rank` column. We can use the `values_drop_na` argument in `pivot_longer()` to drop these rows from the output.

```{r billboard pivot_longer drop_na}
billboard |> 
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE)
```

We can also convert some of the values in the `week` column from character to numeric using the `names_prefix()` argument. This argument allows us to specify a string that will be removed from the values in the `week` column before they are converted to numeric.

```{r mutate billboard}
billboard |> 
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE,
               names_prefix = "wk") |> 
  mutate(week = as.numeric(week))
```


## Widening datasets

`pivot_wider()` is used to widen datasets. It allows us to handle an observation that is spread across *multiple rows*. It takes key-value pairs and spreads them across multiple columns, *decreasing* the number of rows and *increasing* the number of columns. The dataset is split by the `names_from` argument, which specifies the name of the column that contains the names of the new columns. The `values_from` argument specifies the name of the column that contains the values that will be spread across the new columns.

```{r cms_patient_experience}
cms_patient_experience <- cms_patient_experience
cms_patient_experience
```

The unit being studied in the data set is an organization. In this format, though, each organization is spread across 6 rows with 1 row for each of the 6 measurements taken during the survey.

```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
```


To tidy this dataset so that each row is a single observation, we can use `pivot_wider()` to spread the values in the "measure" column across multiple columns.

```{r cms_patient_experience pivot_wider}
cms_patient_experience |> 
  pivot_wider(id_cols = starts_with("org"),   # tells pivot_wider which columns have values that are unique to each row
              names_from = measure_title,
              values_from = prf_rate)
```


# Pivot exercise with Palmer Penguins data

```{r}
penguins <- penguins
penguins
```

We want to make a single plot that shows the distribution of *all* morphometric measurements (`bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`). To plot using `facet_wrap()`, we need to stack the variables into 2 columns: one for measurement name, and one for numerical measurement value. 

```{r lengthen penguins}
penguins_long <- penguins |> 
  pivot_longer(cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
               names_to = "measurement_type",
               values_to = "value")

penguins_long
```

Using `penguins_long` data set, we can plot the distribution of all morphometric measurements using `ggplot` and `facet_wrap()`.

```{r penguins plot with pivot_longer}
penguins_long |> 
  drop_na(value) |> 
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(title = "Morphometric distributions across penguin species",
       x = "Measurement value",
       y = "Frequency") +
  theme(plot.title = element_text(face = "bold")) +
  scale_fill_discrete(name = "Species")

```

We can also widen the `penguins` data set with `pivot_wider()`. This can be useful when you need to present summary statistics as a **cross-tabulated matrix**. 

```{r mass_summary penguins}
mass_summary <- penguins |> 
  drop_na(body_mass_g) |> 
  group_by(species, island) |> 
  summarise(mean_mass = mean(body_mass_g))

mass_summary
```

To create a cross-tabulated matrix that is more appropriate for publication, we can use `pivot_wider()` to spread the `island` variable across multiple columns.

```{r pivot_wider mass_penguins}
mass_matrix <- mass_summary |> 
  pivot_wider(names_from = island,     # which column provides new column headers
              values_from = mean_mass)   # which columns provides values to fill new columns

mass_matrix
# NA values mean no penguins of that species were found on that island
```


# Separating and uniting data tables

```{r}
table3
```

## `separate()`

The `rate` column needs to be split into a `cases` and `population` column. The `separate()` function from the `tidyr` package can be used to split a single column into multiple columns based on a separator. 

```{r separate table3}
table3 |> 
  separate(rate, into = c("cases", "population"))
```

`separate()` defaults to splitting values by any non-alphanumeric character it sees. In `table3`, it split the `rate` column based on the "/" character. If we wanted to split the `rate` column based on a different character, we could use the `sep` argument to specify the separator.

```{r}
table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/")
# gives the same output as the previous code chunk
```

The `cases` and `population` columns need to be converted to numeric types. They are currently character types because this is the default when using `separate()`. 

```{r change column types table 3}
table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/") |> 
  mutate(cases = as.numeric(cases),
         population = as.numeric(population))

# OR

table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```

When you pass an integer to the `sep` argument, `separate()` will split the column at that position. For example, if we wanted to split the `year` column into a `century` and `year` column, we could use `sep = 2` to split the column after the first 2 characters.

```{r separate by integers}
table3 |> 
  separate(year, into = c("century", "year"), sep = 2)
```

## `unite()`

