---
title: "R for Marine Science - Workshop 2"
author: "Katie Dyck"
date: "2026-06-04"
format:
    html:
        embed-resources: true
        code-fold: show
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

# Introduction

This qmd file includes all material covered from the second "R for Marine Science" workshop. The material covered includes:

- Tidying data
- Pivoting wider and longer
- Separating and uniting variables
- Dealing with strings and dates
- Joining and separating data frames/tibbles
- Handling missing values

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

## Housekeeping
```{r Clear session environment}
#| message: true
#| warning: false
rm(list = ls())
```

```{r Connect to GitHub, include=FALSE}
# in the console:
#   credentials::git_credential_ask()
#   usethis::git_sitrep()

# if error message comes up that default branch is origin when it should be "main", enter this in the console:
#   usethis::git_default_branch_rediscover()

# in the console:
#   usethis::use_git()
#   usethis::use_github()

# in RStudio terminal:
  # git remote add origin https://github.com/katiedyck/MB5370_Module02.git
  # git push -u origin main
  # git push
```

## Load packages

```{r load packages}
# Load all required packages for this workshop
library(tidyverse)  # includes `stringr` and `lubridate` packages
library(here)
library(readxl)
library(palmerpenguins)
```

# Tidying data

This is what a tidy data table looks like:

```{r tidy table example}
table1 <- table1
print(table1)
```

This is what an *untidy* data table looks like:

```{r untidy table examples}
table2 <- table2
print(table2)

table3 <- table3
print(table3)
```

A tidy dataset follows these 3 interrelated rules:

- Each variable has its own column
- Each observation has its own row
- Each value has its own cell

The rules are interrelated because only 2 of the 3 rules can be followed at any time.

*All `tidyverse` packages have been designed to work with tidy data.*

# Pivoting data

`pivot_longer()` and `pivot_wider()` are used to reshape data tables. Both functions are made available from the `tidyr` package, which is part of the `tidyverse`.

```{r pivot_longer()}
table1 |> 
  pivot_longer(cols = c("cases", "population"),
               names_to = "case_type",
               values_to = "case_count")
```
```{r pivot_wider}
table2 |> 
  pivot_wider(names_from = "type", values_from = "count")
```


## Lengthening datasets

`pivot_longer()` is used to lengthen datasets. It takes multiple columns and collapses them into key-value pairs, *increasing* the number of rows and *decreasing* the number of columns. The dataset is split by the `names_to` argument, which specifies the name of the new column that will contain the names of the original columns. The `values_to` argument specifies the name of the new column that will contain the values from the original columns.

```{r billboard}
billboard <- billboard
billboard
```

In this dataset, each observation is a song, and the first 3 columns (`artist`, `track`, and `date.entered`) are variables that describe the song. The remaining columns (`wk1` to `wk76`) are not variables, but rather values *of* a variable (week). To tidy this dataset so that each row is a single observation, we can use `pivot_longer()` to collapse the `wk1` to `wk76` columns into key-value pairs.

```{r billboard pivot_longer}
billboard |> 
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               values_to = "rank")
```

As some songs are only on the billboard for a few weeks, there are many `NA` values in the `rank` column. We can use the `values_drop_na` argument in `pivot_longer()` to drop these rows from the output.

```{r billboard pivot_longer drop_na}
billboard |> 
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE)
```

We can also convert some of the values in the `week` column from character to numeric using the `names_prefix()` argument. This argument allows us to specify a string that will be removed from the values in the `week` column before they are converted to numeric.

```{r mutate billboard}
billboard |> 
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE,
               names_prefix = "wk") |> 
  mutate(week = as.numeric(week))
```


## Widening datasets

`pivot_wider()` is used to widen datasets. It allows us to handle an observation that is spread across *multiple rows*. It takes key-value pairs and spreads them across multiple columns, *decreasing* the number of rows and *increasing* the number of columns. The dataset is split by the `names_from` argument, which specifies the name of the column that contains the names of the new columns. The `values_from` argument specifies the name of the column that contains the values that will be spread across the new columns.

```{r cms_patient_experience}
cms_patient_experience <- cms_patient_experience
cms_patient_experience
```

The unit being studied in the data set is an organization. In this format, though, each organization is spread across 6 rows with 1 row for each of the 6 measurements taken during the survey.

```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
```


To tidy this dataset so that each row is a single observation, we can use `pivot_wider()` to spread the values in the "measure" column across multiple columns.

```{r cms_patient_experience pivot_wider}
cms_patient_experience |> 
  pivot_wider(id_cols = starts_with("org"),   # tells pivot_wider which columns have values that are unique to each row
              names_from = measure_title,
              values_from = prf_rate)
```


# Pivot exercise with Palmer Penguins data

```{r}
penguins <- penguins
penguins
```

We want to make a single plot that shows the distribution of *all* morphometric measurements (`bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`). To plot using `facet_wrap()`, we need to stack the variables into 2 columns: one for measurement name, and one for numerical measurement value. 

```{r lengthen penguins}
penguins_long <- penguins |> 
  pivot_longer(cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
               names_to = "measurement_type",
               values_to = "value")

penguins_long
```

Using `penguins_long` data set, we can plot the distribution of all morphometric measurements using `ggplot` and `facet_wrap()`.

```{r penguins plot with pivot_longer}
penguins_long |> 
  drop_na(value) |> 
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(title = "Morphometric distributions across penguin species",
       x = "Measurement value",
       y = "Frequency") +
  theme(plot.title = element_text(face = "bold")) +
  scale_fill_discrete(name = "Species")

```

We can also widen the `penguins` data set with `pivot_wider()`. This can be useful when you need to present summary statistics as a **cross-tabulated matrix**. 

```{r mass_summary penguins}
mass_summary <- penguins |> 
  drop_na(body_mass_g) |> 
  group_by(species, island) |> 
  summarise(mean_mass = mean(body_mass_g))

mass_summary
```

To create a cross-tabulated matrix that is more appropriate for publication, we can use `pivot_wider()` to spread the `island` variable across multiple columns.

```{r pivot_wider mass_penguins}
mass_matrix <- mass_summary |> 
  pivot_wider(names_from = island,     # which column provides new column headers
              values_from = mean_mass)   # which columns provides values to fill new columns

mass_matrix
# NA values mean no penguins of that species were found on that island
```


# Separating and uniting data tables

```{r}
table3
```

## `separate()`

The `rate` column needs to be split into a `cases` and `population` column. The `separate()` function from the `tidyr` package can be used to split a single column into multiple columns based on a separator. 

```{r separate table3}
table3 |> 
  separate(rate, into = c("cases", "population"))
```

`separate()` defaults to splitting values by any non-alphanumeric character it sees. In `table3`, it split the `rate` column based on the "/" character. If we wanted to split the `rate` column based on a different character, we could use the `sep` argument to specify the separator.

```{r}
table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/")
# gives the same output as the previous code chunk
```

The `cases` and `population` columns need to be converted to numeric types. They are currently character types because this is the default when using `separate()`. 

```{r change column types table 3}
table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/") |> 
  mutate(cases = as.numeric(cases),
         population = as.numeric(population))

# OR

table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```

When you pass an integer to the `sep` argument, `separate()` will split the column at that position. For example, if we wanted to split the `year` column into a `century` and `year` column, we could use `sep = 2` to split the column after the first 2 characters.

```{r separate by integers}
table3 |> 
  separate(year, into = c("century", "year"), sep = 2)
```

## `unite()`

`unite()` is the opposite of `separate()`. It combines multiple columns into a single column. The `col` argument specifies the name of the new column that will be created, and the `sep` argument specifies the separator that will be used to combine the values from the original columns.

```{r unite() table5}
table5

table5 |> 
  unite(new, century, year, sep = "")   # no separator - default is "_"
```


# Wrangling strings and data

## `stringr` package

The `stringr` package contains functions for working with strings, all of which begin with "str_".

```{r messy data frame}
# Very messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

print(messy_sites)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    # 4. Remove any trailing or leading underscores
    site_clean = str_replace_all(site_clean, pattern = "^_+|_+$", replacement = ""),
    # 5. Replace underscores with spaces
    site_clean = str_replace_all(site_clean, pattern = "_", replacement = " "),
    # 6. Convert to title case (first letter of each word capitalized)
    site_clean = str_to_title(site_clean)
  )

print(clean_sites)
```

## `lubridate` package

The `lubridate` package allows you to parse dates by spelling out the order of the components (year, month, day, hour, minute, second) in the date format. The functions for parsing dates are `ymd()`, `mdy()`, `dmy()`, `ymd_hms()`, etc.

```{r dates}
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2
```

```{r dates 2}
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)
```


# Joining tables

Sometimes data is split across multiple tables. For example, you might have a massive table of daily fish catch records that only lists a `site_code`, and a separate, smaller metadata table that contains the exact GPS coordinates and zoning status for those site codes. In order to model the data, the tables need to be joined together based on a **key**, or a common column that exists in both tables.

```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

print(observations)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

print(site_metadata)
```

### The workhorses: `left_join()` and `right_join()`

These two functions work the same way but in opposite directions. `left_join()` keeps all rows from the left table and only matching rows from the right table, while `right_join()` keeps all rows from the right table and only matching rows from the left table.

```{r left_join}
joined_data <- observations |> 
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)
```

### The strict filter: `inner_join()`

Only rows that have complete matches in both tables are kept when using `inner_join()`. All other observations are dropped.

```{r inner_join}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))

print(matched_data)
```

### The detective: `anti_join()`

Rather than adding new columns, `anti_join()` filters the left table to *only* show rows that *do not* have a match in the right table. If you try a `left_join()` and end up with unexpected NA values, you can perform an `anti_join()` to find out exactly which site codes are misspelled or missing from your metadata table.

```{r anti_join}
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))

print(missing_context)
```

# Handling missing values

```{r}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)
```
