---
title: "R for Marine Science - Workshop 2"
author: "Katie Dyck"
date: "2026-06-04"
execute:
  echo: false
format:
    html:
        embed-resources: true
        code-fold: show
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

## Introduction

This qmd file includes all material covered from the second "R for Marine Science" workshop. The material covered includes:

- Tidying data
- Pivoting wider and longer
- Separating and uniting variables
- Dealing with strings and dates
- Joining and separating data frames/tibbles
- Handling missing values

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r load packages}
# Load required packages
library(tidyverse)  # includes `stringr` and `lubridate` packages
```

# Tidying data

This is what a tidy data table looks like:

```{r tidy table example}
table1 <- table1
print(table1)
```

This is what an *untidy* data table looks like:

```{r untidy table examples}
table2 <- table2
print(table2)

table3 <- table3
print(table3)
```

A tidy dataset follows these 3 interrelated rules:

- Each variable has its own column
- Each observation has its own row
- Each value has its own cell

The rules are interrelated because only 2 of the 3 rules can be followed at any time.

*All `tidyverse` packages have been designed to work with tidy data.*

# Pivoting data

`pivot_longer()` and `pivot_wider()` are used to reshape data tables. `pivot_longer()` is used to make a wide table longer, while `pivot_wider()` is used to make a long table wider.

```{r pivot_longer()}
table1 |> 
  pivot_longer(cols = c("cases", "population"),
               names_to = "case_type",
               values_to = "case_count")
```
```{r pivot_wider}
table2 |> 
  pivot_wider(names_from = "type", values_from = "count")
```


# Lengthening datasets

