---
title: "R for Marine Science - Workshop 1"
author: "Katie Dyck"
date: "2026-06-03"
execute:
  echo: false
format:
    html:
        embed-resources: true
        code-fold: show
        code-summary: "Show code"
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

I created a new GitHub repository for this module & connected it to RStudio:

```{r Connect to GitHub}
# in the console:
#   credentials::git_credential_ask()
#   usethis::git_sitrep()

# if error message comes up that default branch is origin when it should be "main", enter this in the console:
#   usethis::git_default_branch_rediscover()

# in the console:
#   usethis::use_git()
#   usethis::use_github()

# in RStudio terminal:
  # git remote add origin https://github.com/katiedyck/MB5370_Module02.git
  # git push -u origin main
  # git push
```

Data files (.csv, .txt, .xlsx) were copied from class repository "R4MarineScience" and added to the data folder in my repository for Module 2.

```{r Clear session environment}
## Housekeeping

# Inventory every active object currently residing in session RAM
objects()

# Purge global environment
rm(list = ls())

# Confirm that global session memory is now vacant
objects()
```
# Load necessary packages

```{r Load packages}
library(tidyverse)
library(readxl)
library(palmerpenguins)
```

# Importing different file types

```{r Import data}
# Practice import A
benthic_cover <- read_csv("../data/reef_cover_log.csv")

# Practice import B
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

# Practice import C
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")

# Read in mangrove data
mangrove_data <- read_csv(here::here("data/mangrove_survey_raw.csv"))
mangrove_data

# Use args in read_csv() to read mangrove_data properly
mangrove_data <- read_csv(here::here("data/mangrove_survey_raw.csv"),
                          skip = 5,       # skip the first 5 lines of field notes
                          na = c(".", "NA", "9999", "ND", "blank"))      # convert known text alternatives to true NA
mangrove_data
```

All 4 data sets are tibbles.
You can turn tibbles, which are `tidyverse` data objects, into a data frame:

```{r Covert tibble to data frame}
benthic_cover_df <- as.data.frame(benthic_cover)

print(benthic_cover)
print(benthic_cover_df)
```

Notice the difference between these two data sets. `benthic_cover` tells you its dimensions and variable types (character, double, etc.) below the column headers. 


## Palmer Penguins data set

(The library `palmerpenguins` must be installed and loaded to access the following dataset.)

```{r Load penguin data}

# Load package data into active memory
data("penguins")

# Examine dataset structure
glimpse(penguins)       # tidyverse version
str(penguins)           # base R version

# Generate an exploratory summary matrix
summary(penguins)
```

Note that the `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, `body_mass_g`, and `sex` variables have some NA values.

# Foundational grammar: Slicing, filtering, sorting, and transforming

## `select()` - isolating attributes

Targets variables vertically:

```{r select()}
# Vertically slice specific variables by name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using colon operator
spatial_block <- select(penguins, species:island)
glimpse(spatial_block)

# Discard logistics tracking attributes while keeping everything else using minus sign
clean_scientific_fields <- select(penguins, -year)
glimpse(clean_scientific_fields)
```

## `filter()`

Isolates records (i.e. rows) horizontally based on targeted conditional parameters. R evaluates each row against a logical expression, keeping entries that are `TRUE` and dropping those that are `FALSE` or `NA` (i.e. those that do not meet the expression).

```{r filter()}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Filter out individuals using continuous numerical boundary thresholds
# Keeps only large penguins whose mass exceeds 4500 g
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Keep records that match Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Search for rows matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```


## `arrange()`

Alters how the rows of data are sorted within your data frame, without changing individual cell values

```{r arrange()}
# Sort penguins by ascending body mass (Default setting: smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

## Pipe operator (`|>`)

The pipe operator is a great way to chain multiple operations together. You read from left to right, or top to bottom. You take the data, *then* filter it, *then* mutate it to create a new column, *then* select the columns you need.

The native R pipe syntax is `|>`, but there is another version built into the tidyverse with the magrittr package, which is also commonly used (`%>%`).

Instead of writing:
```{r no pipe operator}
penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")
```

You can just write:

```{r with pipe operator}
penguins_final <- penguins |> 
    mutate(bill_ratio = bill_length_mm / bill_depth_mm) |> 
    filter(species == "Adelie")
```

## `mutate()`

Used to calculate new variables based on ones that already exist in your data.

```{r mutate()}
penguin_ratios <- penguins |> 
    mutate(body_mass_kg = body_mass_g / 1000,
           bill_ratio = bill_length_mm / bill_depth_mm)

glimpse(penguin_ratios)
```


# Data aggregation & ecological summarisation

```{r group_by(), summarise()}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Note that table looks similar, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing groups into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins, 
                                  mean_mass_g = mean(body_mass_g))
print(species_mass_summary)
```

Adelie and Gentoo species data includes NA values - this is an R design feature called the **Missing Value Trap**. Any mathematical function of any kind (sum, mean, sd, etc.) will automatically return NA to prevent miscalculations. This can be overridden by explicitly declaring that missing cells should be dropped during calculations.

```{r missing trap value}
biological_signal <- penguins |> 
    group_by(species, sex) |> 
    summarise(
        sample_size = n(),    # count total number of individuals per category
        mean_mass_g = mean(body_mass_g, na.rm = TRUE),
        sd_mass_g = sd(body_mass_g, na.rm = TRUE)       # standard deviation calculation
    )

print(biological_signal)
```


# Visualizing the `penguins` data

## Visualizing body size and shape

```{r visualizing body size & shape}
penguin_mass_by_island <- penguins |> 
    group_by(species, island) |> 
    summarise(mean_mass_g = mean(body_mass_g, na.rm = TRUE))

print(penguin_mass_by_island)

# Visualize the data
ggplot(penguin_mass_by_island, aes(x = species, y = mean_mass_g)) +
    geom_boxplot() +
    facet_wrap(~ island) +
    labs(x = "Species", y = "Mean Body Mass (g)", title = "Mean Body Mass (g) of Penguin Species found at Palmer Research \nStation, Antarctica") +
    theme(plot.title = element_text(face = "bold"))
```

You can also pipe directly to visualization without creating an intermediate summary table:

```{r piping directly to visualization}
mass_compare_plot <- penguins |> 
    filter(!is.na(body_mass_g)) |>    # remove any NA values from body_mass_g
    group_by(species, island) |> 
    summarize(
        mean_mass = mean(body_mass_g, na.rm = TRUE),
        n = n(),
        sd_mass = sd(body_mass_g, na.rm = TRUE),
        .groups = "drop"
    ) |> 
    ggplot(aes(x = species, y = mean_mass, colour = island)) +
    geom_point(size = 3, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = mean_mass - sd_mass,
                      ymax = mean_mass + sd_mass),
                  width = 0.2,
                  position = position_dodge(width = 0.5)) +
    labs(title = "Mean Body Mass by Species and Island",
         x = "Species",
         y = "Mean Body Mass (g) ± SD ") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))

mass_compare_plot
```

## **Completed morphological plot and summary table**

```{r}
mass_compare_plot2 <- penguins |> 
    filter(!is.na(body_mass_g)) |>    # remove any NA values from body_mass_g
    group_by(species, island) |> 
    summarize(
        mean_mass = mean(body_mass_g, na.rm = TRUE),
        n = n(),
        sd_mass = sd(body_mass_g, na.rm = TRUE),
        se_mass = sd_mass / sqrt(n),
        .groups = "drop"
    ) |> 
    ggplot(aes(x = species, y = mean_mass, colour = island)) +
    geom_point(size = 2, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = mean_mass - se_mass,
                      ymax = mean_mass + se_mass),
                  width = 0.2,
                  position = position_dodge(width = 0.5)) +
    labs(title = "Mean Body Mass by Species and Island",
         x = "Species",
         y = "Mean Body Mass (g) ± SE ") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))

mass_compare_plot2
```

When we compare these two graphs side by side, we see that the standard *error* bars are much smaller than the standard *deviation* bars, which makes sense because the standard error is a measure of how precisely we have estimated the mean, while the standard deviation is a measure of how much variability there is in the data. The standard error will always be smaller than the standard deviation, especially as sample size increases.

```{r}
summary_SD <- penguins |> 
    filter(!is.na(body_mass_g)) |>    # remove any NA values from body_mass_g
    group_by(species, island) |> 
    summarize(
        mean_mass = mean(body_mass_g, na.rm = TRUE),
        n = n(),
        sd_mass = sd(body_mass_g, na.rm = TRUE),
        .groups = "drop"
    )

summary_SE <- penguins |> 
    filter(!is.na(body_mass_g)) |>    # remove any NA values from body_mass_g
    group_by(species, island) |> 
    summarize(
        mean_mass = mean(body_mass_g, na.rm = TRUE),
        n = n(),
        sd_mass = sd(body_mass_g, na.rm = TRUE),
        se_mass = sd_mass / sqrt(n),
        .groups = "drop"
    )

print(summary_SD)
print(summary_SE)
```

Looking at the summary tables, we can see that the standard error decreases as sample size increases (e.g. SE 73.5 for Adelie penguins on Biscoe island (n = 44) vs. SE 45.5 for Gentoo penguins on Biscoe island (n = 68)). These tables also tell us that as sample size (n) increases, so does the mean mass for each species on each island, which makes sense.


# Saving, exporting, and version milestones

```{r Saving tables}
# Create output directories if they do not exist:
if (!dir.exists("outputs/figures")) dir.create("outputs/figures", recursive = TRUE) # folder for figs
if (!dir.exists("outputs/tables")) dir.create("outputs/tables", recursive = TRUE) # folder for tables
if (!dir.exists("Rdata")) dir.create("Rdata", recursive = TRUE) # folder for Rdata objects

# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "outputs/penguin_species_mass_summary.csv")

# 2. Saving our cleaned morphological cohort table as a native R binary file
saveRDS(clean_scientific_fields, "outputs/clean_penguin_morphology_cohort.rds")
```

```{r Saving figures}
ggsave("outputs/mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)

ggsave("outputs/mass_compare_plot2.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)
```
