---
title: "R for Marine Science - Workshop 2 Final Exercise "
author: "Katie Dyck"
date: "2026-06-04"
execute:
  echo: false
format:
    html:
        embed-resources: true
        code-fold: show
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

# Introduction

Here I will be wrangling 4 messy data sets ...

# Housekeeping
```{r Housekeeping}
rm(list = ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r load packages}
# Load necessary packages
library(tidyverse)  # includes stringr, lubridate, ggplot2, dplyr, etc.
library(readxl)     # for reading .xlsx files
```


# Part 1: Loading and Cleaning the data

## Loading the datasets

The dataset *"estuary_catch_log.xlsx"* contains *four sheets* of data that we want to join into **one sheet**.

```{r upload & join fish catch data}
# Use `purrr` package to load "estuary_catch_log.xlsx" file and join all 4 sheets

file_path <- "../data/workshop2/estuary_catch_log.xlsx"
catch_data <- excel_sheets(file_path) |> 
    map_dfr(~ read_excel(path = file_path, sheet = .x))

print(catch_data)

unique(catch_data$site)    # The join was a success! All 4 sheets are now in 1 data table.
```

```{r load remaining datasets}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
print(metadata)

water_quality <- read_csv("../data/workshop2/estuary_sonde_data.csv")
print(water_quality)

dictionary <- read_csv("../data/workshop2/species_dictionary.csv")
print(dictionary)
```


## Cleaning the data

We want the site names to be the same across all data tables.

```{r check data tables}
# Use `unique()` to check for "messy" site and species names columns in the data tables

unique(catch_data$site)     # needs to be cleaned
unique(catch_data$species)  # needs to be cleaned

unique(metadata$site_name)  # needs to be cleaned

unique(water_quality$site)  # needs to be cleaned

unique(dictionary$common_name)  # needs to be cleaned
```

Using the `stringr` package we can standardize the site and species names so they are the same across each data table.

```{r clean catch_data site names}
# Change site names in 'catch_data'

catch_data <- catch_data |> 
    mutate(
        # Replace spaces with underscores
        site = str_replace_all(site, pattern = " ", replacement = "_"),
        # Change text to lowercase
        site = str_to_lower(site)
    )

print(catch_data)
unique(catch_data$site)   # all site names are now in title-case with spaces rather than underscores
```

```{r clean catch_data species names}
# Change species names in 'catch_data'

catch_data <- catch_data |>
    mutate(
        # Change text to title case
        species = str_to_lower(species),
        # Replace spaces between words with underscore
        species = str_replace_all(species, pattern = " ", replacement = "_")
    )

print(catch_data)

unique(catch_data$species)      # species names are now clean
```

```{r clean metadata site names}
# Clean site names in 'metadata'

metadata <- metadata |> 
    mutate(
        # Replace spaces with underscores
        site_name = str_replace_all(site_name, pattern = " ", replacement = "_"),
        # Change text to lowercase
        site_name = str_to_lower(site_name)
    )

print(metadata)

unique(metadata$site_name)    # site names are now clean!
```

```{r clean water_quality site names}
# Clean site names in water_quality

water_quality <- water_quality |> 
    mutate(
        # Replace spaces between words with underscores
        site = str_replace_all(site, pattern = " ", replacement = "_"),
        # Change to lowercase
        site = str_to_lower(site)
    )

print(water_quality)

unique(water_quality$site)    # site names are now clean!
```

```{r clean dictionary common names}
# Clean 'dictionary' common species names

dictionary <- dictionary |> 
    mutate(
        # Replace spaces between words with underscores
        common_name = str_replace_all(common_name, pattern = " ", replacement = "_"),
        # Change to lowercase
        common_name = str_to_lower(common_name)
    )

print(dictionary)

unique(dictionary$common_name)
```

Species names and site names are now standardized across all 4 data tables.


## Parsing Dates & Times

```{r}
glimpse(water_quality)
```

In the water quality data, the `timestamp` column needs to be parsed to a Date-Time object.

```{r}
# View the `timestamp` values
head(water_quality$timestamp, n=20)
```

The `lubridate` package can be used to parse this column from character to Date-Time (`<dttm>`).

```{r parse timestamp column}
# Parse the `timestamp` column

water_quality <- water_quality |> 
    mutate(timestamp = dmy_hm(timestamp))   # current format: DD/MM/YYYY HH:MM

print(water_quality)
```

## Hardware Errors

There are some missing turbidity values in the `water_quality` data table because of occasional turbidity sensor failures that occurred. We need to find them and change them to true NA values so calculations can be performed properly.

```{r}
# Find -999.0 turbidity values in the water_quality data, and change them to true NA values
water_quality |> 
    filter(turbidity == -999.0) |> 
    count()     # there are 567 rows with -999.0 recorded as turbidity value

water_quality <- water_quality |> 
    mutate(turbidity = na_if(turbidity, -999.0))

# Check if -999.0 values were successfully converted to proper NA values
water_quality |> 
    filter(is.na(turbidity)) |> 
    count()     # success!
```


# Part 2

## Summarize water sensor data

