---
title: "R for Marine Science - Workshop 2 Final Exercise"
author: "Katie Dyck"
date: "2026-06-04"
format:
    html:
        embed-resources: true
        code-fold: show
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        number-sections: true
        number-depth: 4
        toc-location: left
        toc-expand: 2
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
---

# Introduction

Here I will be wrangling 4 messy data files containing information on the relationship between water quality and predatory fish species present in the Ross River estuary gradient. All files will be wrangled so that different relationships can be examined:

1. *estuary_catch_log.xlsx*: Field counts of fish species (common names) collected at different locations, spread across multiple sheets.
2. *estuary_metadata.csv*: Spatial coordinates and estuary zones for each site.
3. *estuary_sonde_data.csv*: 
4. *species_dictionary.csv*: 

# Housekeeping
```{r Housekeeping}
#| message: true
#| warning: false
rm(list = ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r Connect to GitHub, include=FALSE}
# in the console:
#   credentials::git_credential_ask()
#   usethis::git_sitrep()

# if error message comes up that default branch is origin when it should be "main", enter this in the console:
#   usethis::git_default_branch_rediscover()

# in the console:
#   usethis::use_git()
#   usethis::use_github()

# in RStudio terminal:
  # git remote add origin https://github.com/katiedyck/MB5370_Module02.git
  # git push -u origin main
  # git push
```

# Packages
```{r load packages, eval=FALSE}
library(tidyverse)  # includes stringr, lubridate, ggplot2, dplyr, purrr, etc.
library(readxl)     # for reading .xlsx files
```

# Loading and Cleaning the data

## Load files

The catch data is divided across 4 sheets that we want to join together into one.

```{r upload & join fish catch data, echo=FALSE}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)


# Use `purrr` package to load "estuary_catch_log.xlsx" file and join all 4 sheets
# file_path <- "../data/workshop2/estuary_catch_log.xlsx"
# catch_data <- excel_sheets(file_path) |>
#     map_dfr(~read_excel(path = file_path, sheet = .x))
```

## Inspect individual sheets
```{r investigate indiv catch data sheets}
print(catch1)
print(catch2)
print(catch3)
print(catch4)
```

## Bind rows together
```{r bind catch data rows}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4)
```

## Examine catch data
```{r Examine catch data}
print(catch_data)
```

We want to know how many unique levels of `site` and `species` we have; we can't tell how many there are from these outputs. One option is to check these columns explicitly using `distinct()`:

```{r unique species and sites}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

We have 4 species as well as 4 sites which seem to be located at different points along the Ross River. We need the site names to have a uniform format; we can use the `stringr` package to make `site` names consistent so we can then use the taxonomic dictionary to translate to scientific names.


## Load metadata

```{r load remaining datasets, echo=FALSE}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
glimpse(metadata)
```

## Clean metadata

```{r check data tables}
metadata |> 
  distinct(site_name)
```

Site names in `metadata` are all over the place. We need to standardize them.

```{r clean catch_data site names}
catch_data <- catch_data |> 
  mutate(
        # Replace spaces with underscores
        site = str_replace_all(site, pattern = " ", replacement = "_"),
        # Change text to lowercase
        site = str_to_lower(site)
    ) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor))
```

```{r clean metadata site names}
metadata <- metadata |> 
  mutate(
        # Replace spaces with underscores
        site_name = str_replace_all(site_name, pattern = " ", replacement = "_"),
        # Change text to lowercase
        site_name = str_to_lower(site_name)
    ) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor))
```

## Recheck:

```{r recheck}
catch_data |> 
  distinct(site)

metadata |> 
  distinct(site_name)
```

## Load taxonomic dictionary data

```{r load dictionary data}
dictionary <- read_csv("../data/workshop2/species_dictionary.csv")
glimpse(dictionary)
```

## Wrangle sensor data

```{r load sensor data, echo=FALSE}
sensor_raw <- read_csv("../data/workshop2/estuary_sonde_data.csv")
str(sensor_raw)
```

### Notes
- The site names look good; they match the names from the fish catch dataset.
- Environmental variables need cleaning: some negative salinity values, some -999 turbidity values (failed values)
- `timestamp` is a character-type column. We will probably want to parse this to a date or date-time column.

### Clean sensor data

```{r clean sensory data}
sensor_data <- sensor_raw |> 
    mutate(
      # Parse the character string to Day/Month/Year Hour:Minute
      datetime = dmy_hm(timestamp),
      # Chance -999.0 to NA values
      turbidity = na_if(turbidity, -999),
      # Convert negative salinity values to NA
      salinity = ifelse(salinity < 0, NA, salinity)
    )
```

### Summarize 15-minute sensor data into daily means

```{r}
sensor_daily <- sensor_data |>
  mutate(date = as_date(floor_date(datetime, "day"))) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop")
```

## Standardize catch species names to match dictionary

```{r}
dictionary |> 
  distinct(common_name)

catch_data |> 
  distinct(species)
```

## Standardize variable in catch log to match dictionaries
```{r clean catch species names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

## Check common names in catch log match those in the taxonomic dictionary
```{r}
catch_data |> 
  distinct(species)
dictionary |>
  distinct(common_name)
```


# Taxonomic Translation (Common to Scientific)

```{r taxonomic translation}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(dictionary, by = join_by(common_name))

print(catch_data)
```

## Need to:

1. Join with `metadata` and `sensor_daily` data
2. Deal with implicit zero observations
  - 4 x sp
  - 4 x sites
  - 30 surveys

4*4*30

## join catch data with metadata and sensor data
```{r join catch data with metadata and sensor data}
catch_data_final <- catch_data |> 
  left_join(metadata, by = join_by(site == site_name)) |> 
  left_join(sensor_daily, by = join_by(site, date))

print(catch_data_final)
```

# Zero-catch framework
```{r built zero-catch framework}
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity), scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```
First check there is an actual difference in salinity among zones:
```{r mean/sd/se salinity}
# sal_sum
catch_data_final |> 
  group_by(scientific_name, zone) |> 
  summarise(
    # se_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# Or with plotrix
library(plotrix)
catch_data_final |> 
  group_by(scientific_name, zone) |> 
  summarise(
    se_salinity = std.error(count),
    # sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
```

