---
title: "R4MS - Workshop 2 Final Exercise"
author: "Katie Dyck"
date: "2026-06-04"
format:
    html:
        embed-resources: true
        code-fold: show
        code-tools: true
        code-block-bg: true
        code-block-border-left: "darkgrey"
        code-overflow: wrap
        toc: true
        toc-location: left
        toc-expand: 2
        number-sections: true
        number-depth: 4
        toc-title: Contents
        syntax-highlighting: github
        theme: cosmo
        fig-height: 6
        fig-width: 10
        fig-cap-location: bottom
editor_options: 
    chunk_output_type: inline
---

# Introduction

Here we will be wrangling messy data files on water quality and predatory fish species from the Ross River estuary fish surveys. We know that there are 4 survey sites, 4 species, and 30 days of sampling. One of the files contains metadata on the sites and sensor data on the environmental conditions at each site.

# Housekeeping
```{r Housekeeping}
#| message: true
#| warning: false
rm(list = ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r Connect to GitHub, include=FALSE}
# in the console:
#   credentials::git_credential_ask()
#   usethis::git_sitrep()

# if error message comes up that default branch is origin when it should be "main", enter this in the console:
#   usethis::git_default_branch_rediscover()

# in the console:
#   usethis::use_git()
#   usethis::use_github()

# in RStudio terminal:
  # git remote add origin https://github.com/katiedyck/MB5370_Module02.git
  # git push -u origin main
  # git push
```

# Packages
```{r packages}
library(tidyverse)  # includes stringr, lubridate, ggplot2, dplyr, purrr, etc.
library(readxl)     # for reading .xlsx files
```

# Loading and Cleaning the data

## Load catch data

The catch data is divided across 4 sheets in one excel file.

Load sheets individually:

```{r load catch data}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

```{r load catch day V2, echo=FALSE}
# Use `purrr` package to load "estuary_catch_log.xlsx" file and join all 4 sheets
# file_path <- "../data/workshop2/estuary_catch_log.xlsx"
# catch_data <- excel_sheets(file_path) |>
#     map_dfr(~read_excel(path = file_path, sheet = .x))
```

## Inspect individual sheets

```{r investigate catch data sheets}
#| code-fold: true

print(catch1)
print(catch2)
print(catch3)
print(catch4)
```
There are variations in the species and site names, but they otherwise look to be the same. We can clean after binding all the rows together.

## Bind rows together

```{r bind catch data rows}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4)
```

## Examine catch data

```{r Examine catch data}
print(catch_data)
```

We want to know how many levels of `site` and `species` we have, which we can't tell how many there are from these outputs. One option is to check these columns explicitly using `distinct()`:

```{r unique species and sites}
#| code-fold: true

catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

### Next steps for catch data

We need the site names to have a uniform format. It looks like sites correspond to unique zones on the Ross River, but we need to load the metadata to confirm this.
We can use the `stringr` package to make `site` names consistent so we can then load the taxonomic dictionary file to translate common to scientific species names.

## Load metadata

```{r load remaining datasets, echo=FALSE}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
glimpse(metadata)
```

## Clean metadata

```{r check data tables}
#| code-fold: true

metadata |> 
  distinct(site_name)
```

Site names in `metadata` are all over the place. We need to standardize them so we can join this table with the catch data table.

### Standardize site names in catch data and metadata

Aside from simply standardizing the site names, we also want to convert the character variables to factors.

```{r clean catch_data site names}
catch_data <- catch_data |> 
  mutate(
        # Replace spaces with underscores
        site = str_replace_all(site, pattern = " ", replacement = "_"),
        # Change text to lowercase
        site = str_to_lower(site)
    ) |> 
  # Convert all character to factor
  mutate(across(where(is.character), as_factor))
```

```{r clean metadata site names}
metadata <- metadata |> 
  mutate(
        # Replace spaces with underscores
        site_name = str_replace_all(site_name, pattern = " ", replacement = "_"),
        # Change text to lowercase
        site_name = str_to_lower(site_name)
    ) |> 
  # Convert all character to factor
  mutate(across(where(is.character), as_factor))
```

### Recheck site names

```{r recheck site names}
#| code-fold: true

catch_data |> 
  distinct(site)

metadata |> 
  distinct(site_name)
```

Site names are now standardized across the catch data and metadata files.

## Load taxonomic dictionary

```{r load dictionary data, echo=FALSE}
dictionary <- read_csv("../data/workshop2/species_dictionary.csv")
glimpse(dictionary)
```
## Standardize species (common) names in catch data

The common species names are already in a standardized format. We need to standardize the species names in the catch data to match. 

```{r fix catch data species names}
#| code-fold: true

catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, pattern = " ", replacement = "_")
  )

catch_data |> 
  distinct(species)
```

### Create new variable ("scientific_name") in catch log by joining with taxonomic dictionary 
```{r join-species-dict}
#| code-fold: true

catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(dictionary, by = join_by(common_name))
```

```{r}
glimpse(catch_data)
```

Looks good - the species and site names are fixed.

## Load and wrangle sensor data

```{r load sensor data, echo=FALSE}
sensor_raw <- read_csv("../data/workshop2/estuary_sonde_data.csv")
str(sensor_raw)
```

- Sonde data was collected every 15 minutes for the same 4 sites across the 30 days of sampling. The data was sourced in a daily format, and outliers were removed.
  - `timestamp` is a character-type column. We will probably want to parse this to a date or date-time column.
- There are three different environmental variables measured: *temperature*, *salinity*, and *turbidity*.
  - Environmental variables need cleaning: some negative salinity values, some -999 turbidity values (failed values).
- **Could investigate effects of all three on fish spp abundance or pick one? Need to think about prior research and hypotheses etc.**


## Wrangle sensor data

```{r clean sensor data}
#| code-fold: true

sensor_data <- sensor_raw |> 
    mutate(
      # Parse the character string to Day/Month/Year Hour:Minute
      datetime = dmy_hm(timestamp),
      # Chance -999.0 to NA values
      turbidity = na_if(turbidity, -999),
      # Convert negative salinity values to NA
      salinity = ifelse(salinity < 0, NA, salinity)
    )

print(sensor_data)
```

### Summarize 15-minute sensor data into daily means

```{r summ 15min sensor data}
#| code-fold: true

sensor_daily <- sensor_data |>
  mutate(date = as_date(floor_date(datetime, "day"))) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop")

print(sensor_daily)
```

# Join catch data with metadata and sensor data

```{r join catch data with metadata and sensor data}
catch_data_final <- catch_data |> 
  left_join(metadata, by = join_by(site == site_name)) |> 
  left_join(sensor_daily, by = join_by(site, date))

print(catch_data_final)
```

Note that we only have 427 rows of data...4 species, 4 sites, and 30 surveys = 4 * 4 * 30 = **480**! There should be 480 rows in our data table.

Some days there were no species counted at some sites, so these rows were omitted from the data. We need to wrangle the data so that zero counts are represented as 0 rather than NA. This is important when it comes time to data visualization because we want to be able to distinguish between true absenses (0) and missing data (NA).

# Wrangle final dataset

```{r built zero-catch framework}
#| code-fold: true

catch_data_final <- catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity), scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )

print(catch_data_final)
```

**WE HAVE SUCCESSFULLY WRANGLED THE DATA!**

# Analyzing the Data

We first want to check if there is an actual difference in the environmental variables among zones:

## Salinity

```{r salinity summ}
#| code-fold: true

sal_sum <- catch_data_final |> 
  group_by(zone) |> 
  summarise(
    # se_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

print(sal_sum)

# Or with plotrix
# library(plotrix)
# catch_data_final |> 
#   group_by(scientific_name, zone) |> 
#   summarise(
#     se_salinity = std.error(count),
#     # sd_salinity = sd(mean_salinity, na.rm = TRUE),
#     median_salinity = median(mean_salinity, na.rm = TRUE),
#     .groups = "drop"
#   )
```

```{r salinity summ plot}
#| code-fold: true

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)") +
  theme(plot.title = element_text(face = "bold"))
```

As expected, there are significant differences in salinity between zones of the Ross River. We also know there are differences between species in terms of their response to salinity gradients.

## Turbidity
```{r turbidity plot}
#| code-fold: true

turb_sum <- catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Turbidity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Turbidity (NTU)") +
  theme(plot.title = element_text(face = "bold"))
```

## Temperature

```{r temperature plot}
#| code-fold: true

temp_sum <- catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), 
                width = 0.2) +
  theme_minimal() +
  labs(title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)") +
  theme(plot.title = element_text(face = "bold"))
```

We'll choose salinity as our main environmental variable of interest, but we know that there are also differences in turbidity and temperature across the zones, which could also be influencing fish abundance patterns. **We can discuss this in the interpretation of our results and consider potential confounding effects of these other environmental variables in our discussion.**

## Remake salinity plot nicely for publication
```{r}
#| code-fold: true
#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```

## Patterns in Fish Abundance

### Calculate summary stats 
```{r}
#| code-fold: true

fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )

print(fish_stats)
```

### Summary stats plot
```{r}
#| code-fold: true

fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(data = catch_data_final, aes(x = zone, y = count), 
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) +
  
  geom_point(position = "dodge", size = 4, pch = 23) +
  
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
    legend.position = "none"
  ) +
  labs(
    title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    caption = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
    )
```